---
title: "Bayesian Gompertz model that handles both truncation and covariates"
author: Maria O
---

# Imports
```{r, eval = F}
library(rstan)
library(flexsurv)
library(tidyverse)
library(bayesplot)
library(rstanarm)
library(gridExtra)
library(cowplot)

source("../gompertz_functions.R")
options(mc.cores = 4)
```


# How the model works
We assume a proportional hazards model. Each person $i$ has $k$ predictor variables

$$
h_i(x) = \alpha \times \exp\bigg(\beta x + \sum_{j=1}^k b_j c_j \bigg)
$$
where

$$
\alpha \times \exp{\beta x_i} \ \text{ is the baseline Gompertzian hazard at age } x \\
\exp\bigg(\sum_{j=1}^k b_j c_j \bigg) \ \text{are the effects of each covariate} \ c_j \ \text{mutliplied by coefficient} \ b_j
$$

Input data: An vector $\mathbf{x}$ of $N$ observed death ages, an $N$ by $K$ matrix of covariates, two constants: U and L denoting upper and lower age limits. (All one-dimensional vectors are actually column vectors to Stan, so I am writing their transposes.)

\begin{equation}
L, U \in \mathbb{Z}^+ \\

\mathbf{x}^T = [x_1, x_2, .... , x_n] \\

\textbf{C} =
\begin{bmatrix}
    c_{11}       & c_{12}  & \dots & c_{1k} \\
    c_{21}       & c_{22}  & \dots & c_{2k} \\
    \ \vdots & \vdots  & \ddots & \vdots \\
    c_{n1}       & c_{n2}  & \dots & c_{nk}
\end{bmatrix} \\
\end{equation}


And these are the parameters we estimate:

$$
\mathbf{b}^T = [ b_1, b_2, ... b_k ] \\
M \ \text{(modal age at death in baseline Gompertz schedule)} \\
\beta \ \text{(mortality parameter in baseline Gompertz schedule)}
$$

Notes that what we get OUT of the model is parameterized with $M$ rather than $\alpha$. The internal workings of the model compute $\alpha$ but the user only has to worry about $M$, as it is easier to put a prior on $M$.(is it possible to parameterize everything in terms of the mode rather than alpha? would be less computationally expensive...)

Compute the likelihoods...

$$
h(x_i|\alpha, \beta, \mathbf{b} )= \alpha \times \exp (\beta x_i) \times \exp (\textbf{C}_{i} \cdot \mathbf{b} )
$$

for the purpose of typing fewer things define:

$$
q := \alpha \times \exp (\textbf{C}_{i} \cdot \vec{b} ) \\
h(x_i|q, \beta)= q \times \exp (\beta x_i)
$$

Additionally, we presume that the data observed are only gathered from part of the distribution, bounded by ages L and U.

$$
\mathcal{L}(q, \beta |x) = \prod_{i=1}^N {f(x_i|q,\beta) \over F(U|q,\beta)-F(L|q,\beta)} \\ 



\log \mathcal{L}(q, \beta |x) = \sum_{i=1}^N \log \bigg[{f(x_i|q,\beta) \over F(x_{max}|q,\beta)-F(x_{min}|q,\beta)}\bigg]
$$

where $f(x_i)$ is the PDF of the death distribution:
$$
f(x_i|q,\beta) = q \times \exp\bigg( {q \over \beta}\big(\exp{(\beta x_i)}-1\big)    \bigg)
$$
and $F(x_i)$ the CDF:
$$
F(x_i|q,\beta)= 1- \exp\bigg(- {q \over \beta}\big(\exp{(\beta x_i)}-1\big)    \bigg)
$$
Since $x_{min}$ and $x_{max}$ are set, 
$$
F(U|q,\beta)-F(L|q,\beta) = \exp \bigg( -{q \over \beta} (e^{\beta L}-1) \bigg) - \exp \bigg( -{q \over \beta}(e^{\beta U}-1) \bigg)
$$
*(holy inconsistent notation, batman)*

# Model text
```{r, eval = F}
trunc_mv_model_text = "
functions {
  real get_alpha(real mode, real beta) {
    real alpha = beta/exp(mode*beta);
    return(alpha);
  }
  
  real get_pgomp_between(int U, int L, real mutliplier, real beta) {
    real denom = exp((-mutliplier/beta) * (exp(beta*L) -1 )) - exp((-mutliplier/beta) * (exp(beta*U) -1 ));
    return(denom);
  }

  real gompertz_log(vector x, matrix c, int U, int L, real mode, real beta, vector b){
    real alpha = get_alpha(mode, beta);
    int len = num_elements(x);
    vector[len] probs;
    real multiplier;
    real num;
    real denom;
    real log_like;
    for (i in 1:len) {
      multiplier = alpha * exp(dot_product(c[i], b));
      num = multiplier * exp(beta*x[i]) * exp((-multiplier/beta) * (exp(beta*x[i])-1)); 
      denom = get_pgomp_between(U, L, multiplier, beta);
      probs[i] = num/denom;
    }
    log_like = sum(log(probs));
    return log_like;
  }
}

data {
  int N;           // number of observations
  int K;           // number of predictors
  int U;           // upper age of death observed
  int L;           // lowest age of death observed
  matrix[N,K] c;   // predictor matrix
  vector[N] x;     // observed ages at death
}

parameters {
  real<lower=0> mode;       // modal age at death param
  real<lower=0> beta;       // 'force of mortality' param
  vector[K] b;              // coefficients
}

model {
  x ~ gompertz(c, U, L, mode, beta, b);     // likelihood of observing this age at death, given parameters
  mode ~ uniform(30, 100);                   // prior for mode
  beta ~ uniform(0.001, 0.3);               // prior for beta
  b ~ uniform(-2,2);                        // priors for b vector -- all the same for simplicity
}
"

trunc_mv_gomp_model = stan_model(model_code = trunc_mv_model_text,
                   model_name='gompertz with truncation, covariates')
```

do i actually need to declare all those varaiables before they are computed in the _log function??


# Trying it out on some fake data
using one nominal and one continuous variable.

```{r, eval = F}
set.seed(50)
n <- 1000

# real baseline gompertz parameters
beta_actual <-  1/12
mode_actual <- 70
alpha_actual <- getAlpha(M=mode_actual, beta=beta_actual)

# fake covariates
cont_var_vec <- rnorm(n=n) #contiunous, random from normal(0,1)
b_cont_actual <- 0.5
nom_var_vec <- rbinom(n=n, size=1, prob=0.5) #dummy variable, assigned from binomial with p=50% 
b_nom_actual <- -0.4

# generate death ages
death_ages_with_cv <- rgompertz(n=n,
                                shape=beta_actual,
                                rate=alpha_actual*exp(b_cont_actual*cont_var_vec)*exp(b_nom_actual*nom_var_vec) )

# add truncation to get *observed* deaths
lowest_age <- 65
highest_age <- 95
death_ages_with_cv_trunc <- death_ages_with_cv[death_ages_with_cv >= lowest_age & death_ages_with_cv <= highest_age]

hist(death_ages_with_cv, col='gray', border='darkgray', breaks=seq(0,110,1))
hist(death_ages_with_cv_trunc, col='red', border='darkgray', breaks=seq(0,110,1), add=T)
curve(n*alpha_actual * exp(beta_actual*x) * exp((-alpha_actual/beta_actual) * (exp(beta_actual*x)-1)), 
      from=0, to=110, add=TRUE, col = 'darkblue', lty=2, lwd=2)
```
Dashed Line: Baseline Gompertz(alpha, beta) distribution. What we should see if covariates are all zeroed.
Gray histogram: Distribution of deaths when convariates are factored in.
Red histogram: The Deaths we "observe", restricted to a certain window of ages. Only these go into model as data.



# Imitate the data frame format that these variables would be in
nominal covariates should be turned into dummies with reference category dropped

```{r, eval = F}
# get the covariates only for the observed death ages
cont_var_vec_trunc <- cont_var_vec[death_ages_with_cv >= lowest_age & death_ages_with_cv <= highest_age]
nom_var_vec_trunc <- nom_var_vec[death_ages_with_cv >= lowest_age & death_ages_with_cv <= highest_age]

# put them in a matrix
cov_matrix <- matrix(ncol=2, data=c(cont_var_vec_trunc, nom_var_vec_trunc))
head(cov_matrix)
```



#Initialize and run

```{r, eval = F}
# list the data
trunc_mv_stan_data = list(x = death_ages_with_cv_trunc,
                          c = cov_matrix,
                          N = length(death_ages_with_cv_trunc),
                          K = ncol(cov_matrix),
                          U = highest_age,
                          L = lowest_age)

# initialize parameters
init_gomp_trunc_mv <- function () {list(beta=0.06, mode=50, b=c(0,0))} 

trunc_mv_fit =  sampling(object = trunc_mv_gomp_model, 
                 data = trunc_mv_stan_data, 
                 init = init_gomp_trunc_mv,
                 chains = 2,
                 iter   = 1000, 
                 warmup = 200)

```

# Posterior stuff -- how close did I get?

```{r, eval = F}
traceplot(trunc_mv_fit)
```
the chains converged! how close are the parameter estimates?

```{r, eval = F}
print(trunc_mv_fit, digits=4)
```


```{r, message=F, eval = F}
h1 <- stan_hist(trunc_mv_fit, pars='mode', fill='coral', color='coral2') +
  geom_vline(xintercept=mode_actual, linetype='dotdash')
h2 <- stan_hist(trunc_mv_fit, pars='beta', fill='dodgerblue3', color='dodgerblue4') +
  geom_vline(xintercept=beta_actual, linetype='dotdash')
h3 <- stan_hist(trunc_mv_fit, pars='b[1]', fill='maroon3', color='maroon4') +
  geom_vline(xintercept=b_cont_actual, linetype='dotdash')
h4 <- stan_hist(trunc_mv_fit, pars='b[2]', fill='maroon3', color='maroon4') +
  geom_vline(xintercept=b_nom_actual, linetype='dotdash')

grid.arrange(h1, h2, h3, h4, nrow = 2)
```

not the best prediction for b_2. Wonder if that's a problem with the model or just randomness (either in the MCMC draws or generating the initial distribution of deaths)?

```{r, eval = F}
set.seed(50)
n <- 1000

## function to generate data 

generate_truncated_data <- function(n = 1000){
  # real baseline gompertz parameters
  beta_actual <-  1/12
  mode_actual <- 70
  alpha_actual <- getAlpha(M=mode_actual, beta=beta_actual)
  
  # fake covariates
  cont_var_vec <- rnorm(n=n) #contiunous, random from normal(0,1)
  b_cont_actual <- 0.5
  nom_var_vec <- rbinom(n=n, size=1, prob=0.5) #dummy variable, assigned from binomial with p=50% 
  b_nom_actual <- -0.4
  
  # generate death ages
  death_ages_with_cv <- rgompertz(n=n,
                                  shape=beta_actual,
                                  rate=alpha_actual*exp(b_cont_actual*cont_var_vec)*exp(b_nom_actual*nom_var_vec) )
  
  # add truncation to get *observed* deaths
  lowest_age <- 65
  highest_age <- 95
  death_ages_with_cv_trunc <- death_ages_with_cv[death_ages_with_cv >= lowest_age & death_ages_with_cv <= highest_age]
  
  # get the covariates only for the observed death ages
  cont_var_vec_trunc <- cont_var_vec[death_ages_with_cv >= lowest_age & death_ages_with_cv <= highest_age]
  nom_var_vec_trunc <- nom_var_vec[death_ages_with_cv >= lowest_age & death_ages_with_cv <= highest_age]
  
  # put them in a matrix
  cov_matrix <- matrix(ncol=2, data=c(cont_var_vec_trunc, nom_var_vec_trunc))
  
  tuncated_data_and_covariates <- list(death_ages_with_cv_trunc, cov_matrix)
  
  return(tuncated_data_and_covariates)
}

generate_truncated_data(n = 1000)
```

```{r, eval = F}
truncated_results <- list()

for (i in 1:10) {

tuncated_data_and_covariates <- generate_truncated_data(n = 1000) 

death_ages_with_cv_trunc <- tuncated_data_and_covariates[[1]]
cov_matrix <- tuncated_data_and_covariates[[2]]

# list the data
trunc_mv_stan_data = list(x = death_ages_with_cv_trunc,
                          c = cov_matrix,
                          N = length(death_ages_with_cv_trunc),
                          K = ncol(cov_matrix),
                          U = highest_age,
                          L = lowest_age)

# initialize parameters
init_gomp_trunc_mv <- function () {list(beta=0.06, mode=65, b=c(0,0))} 

trunc_mv_fit =  sampling(object = trunc_mv_gomp_model, 
                 data = trunc_mv_stan_data, 
                 init = init_gomp_trunc_mv,
                 chains = 2,
                 iter   = 1000, 
                 warmup = 200,
                 refresh=0)

mode <- mean(rstan::extract(trunc_mv_fit)$mode)
beta <- mean(rstan::extract(trunc_mv_fit)$beta)
b_1 <- mean(rstan::extract(trunc_mv_fit)$b[,1])
b_2 <- mean(rstan::extract(trunc_mv_fit)$b[,2])

cat("iteration = ", i, "\n")

truncated_results[[i]] <- tibble(mode, beta, b_1, b_2)
}

truncated_results_df <- bind_rows(truncated_results)

truncated_results_df %>% 
  summarize_all(mean)

truncated_results_df
```

```{r, eval = F}
results_long <- truncated_results_df %>%
  mutate(id = row_number()) %>% 
  pivot_longer(cols = -id) %>% 
  mutate(true = case_when(
    name == "mode" ~70,
    name == "beta" ~ 1/12,
    name == "b_1" ~ .5,
    name == "b_2" ~ -.4,
  )) %>% 
  group_by(name) %>% 
  mutate(mean = mean(value)) %>% 
  ungroup()

truncated_covariate_plot <- results_long %>% 
  ggplot() +
  geom_histogram(aes(x = value), fill = "grey", color = "black")+ 
  theme_bw(20) +
  geom_vline(aes(xintercept = true, colour="True"), size = 1) + 
  geom_vline(aes(xintercept = mean, color='Estimated'), size = 1) + 
  labs(title = "Simulation Results",
       subtitle = "R = 300; N = 1,000",
       x = "Mode",
       y = "Count") +
  scale_color_manual(name = "", values = c("True" = "blue", "Estimated" = "red")) + 
  theme(legend.position = "bottom") + 
  facet_wrap(~name, scales = "free", nrow = 2)

ggsave(plot = truncated_covariate_plot, filename = "figures/simulation_truncation_covariates.png", width = 12, height = 6)

```


