---
title: "Simple Gompertzian Mortality Model with Truncation"
author: "Maria Osborne / Leslie Root / Dennis Feehan"
---

# Imports

```{r}
library(rstan)
library(flexsurv)
library(tidyverse)
library(here)
library(janitor)
library(gghighlight)

source('../gompertz_functions.R')
```


# Model Set-up

```{r}
simple_model_text = "
functions {
  real get_alpha(real mode, real beta) {
    real alpha = beta/exp(mode*beta);
    return(alpha);
  }

  real gompertz_lpdf(real x, real mode, real beta){
    real alpha = get_alpha(mode, beta);
    real prob = alpha * exp(beta*x) * exp((-alpha/beta) * (exp(beta*x)-1));
    return log(prob);
  }
 
  real gompertz_lcdf(real x, real mode, real beta){
    real alpha = get_alpha(mode, beta);
    real cdf = 1 - exp((-alpha/beta)*(exp(beta*x)-1));
    return log(cdf);
  }

  real gompertz_lccdf(real x, real mode, real beta){
    real alpha = get_alpha(mode, beta);
    real ccdf = exp((-alpha/beta)*(exp(beta*x)-1));
    return log(ccdf);
  }
 
}

data {
  int N;         // number of observations
  real L;       // lower truncation bound
  real U;       // upper truncation bound
  real<lower=L,upper=U> x[N];   // observed ages at death
}

parameters {
  real<lower=0> mode;       // modal age at death param
  real<lower=0.001> beta;   // 'force of mortality' param
}

model {
    for (i in 1:N) {
        x[i] ~ gompertz(mode, beta) T[L,U];
       
      }  // likelihood of observing this age at death, given mode and beta <-- modified
  mode ~ uniform(60, 90);        // prior for m
  beta ~ uniform(0.001, 0.3);    // prior for beta
}
"
# Initialize the parameters to some reasonable values, otherwise
# stan may attempt to take log likelihood of non-positive values
init_gomp <- function () {list(beta=0.05, mode=75)} 

# compiling the model takes minute or two
simple_gomp_model = stan_model(model_code = simple_model_text,
                   model_name='gomp')
```


# Generated data
```{r}
set.seed(36)

## set number of simulatons 
n <- 10000

## set gompertz parameters 
beta_actual <-  1/12
mode_actual <- 85
alpha_actual <- getAlpha(M=mode_actual, beta=beta_actual)

## set truncation 
truncation_lower <- 65
truncation_upper <- 100

## generate data 
death_ages <- rgomp_mode(n = n, M = mode_actual, beta = beta_actual)

## truncacated data
death_ages_truncated <- subset(death_ages, death_ages >= truncation_lower & death_ages <= truncation_upper)

# visualization to make sure I got the PDF correct...
tibble(death_ages) %>% 
  mutate(window = case_when(
    death_ages > truncation_lower & death_ages < truncation_upper ~ "Y",
    TRUE ~ "N"
  )) %>% 
ggplot(aes(x = death_ages, fill = window)) + 
  geom_histogram(aes(y=..count../sum(..count..)),
                 color = "black",
                 binwidth = 1) + 
  stat_function(fun=function(x) alpha_actual * exp(beta_actual*x) * exp((-alpha_actual/beta_actual) * (exp(beta_actual*x)-1)), color = "blue", size = 1) +
  theme_classic(15) + 
  scale_fill_manual(values = c("grey", "red")) + 
  labs(x = "Death Age",
       y = "Density") + 
  theme(legend.position="none")
```


# Run the model (can take a couple minutes depending on how big N is)
```{r}
simple_stan_data = list(x = death_ages_truncated, N = length(death_ages_truncated), L = 65, U = 100)

# Generate the stan fit object
simple_fit =  sampling(object = simple_gomp_model, 
                 data = simple_stan_data, 
                 init = init_gomp,
                 chains = 2,
                 iter   = 1000, 
                 warmup = 200)
```



# Posterior checks
```{r}
# summary, credible intervals
print(simple_fit, probs= c(0.025, 0.5, .975), digits=3)

# make sure the chains converged
traceplot(simple_fit)

# quick visual check to make sure the real parameters are in the credible interval
stan_hist(simple_fit, pars='mode', fill='coral', color='darkgray') +
  geom_vline(xintercept=mode_actual, linetype='dotdash')
stan_hist(simple_fit, pars='beta', fill='lightblue', color='darkgray') +
  geom_vline(xintercept=beta_actual, linetype='dotdash')
```

# HMD data
Could try with any HMD data you want, but I did it with one year of US females and it took 4 hours to run, so. This is Australia, much smaller population.
```{r}
deaths <- read_table(here("stan/data/Deaths_1x1_AU.txt"), skip = 2)

deaths <- deaths %>% 
  clean_names() %>% 
  mutate(age = as.numeric(ifelse(age=="110+", "110", age))) %>% 
  filter(age<=100) %>%
  filter(age>=65) %>%
  pivot_longer(female:total, names_to = "sex", values_to = "deaths")
  
deaths <- deaths %>% 
  filter(year==1989, sex == "female") %>% 
  mutate(deaths = as.integer(deaths))

hmd_deaths_truncated <- NULL
for (i in 1:nrow(deaths)) {
  hmd_deaths_truncated <- c(hmd_deaths_truncated,rep(as.numeric(deaths[i,2]),as.numeric(deaths[i,4])))
}

hist(hmd_deaths_truncated, breaks=seq(40,115,1), freq=FALSE, col = 'gray', border='darkgray')
```

```{r}
hmd_stan_data = list(x = hmd_deaths_truncated, N = length(hmd_deaths_truncated), L = 65, U = 100)

# Generate the stan fit object
hmd_fit =  sampling(object = simple_gomp_model, 
                 data = hmd_stan_data, 
                 init = init_gomp,
                 chains = 2,
                 iter   = 1000, 
                 warmup = 200)
```

# Posterior checks
```{r}
# summary, credible intervals
print(hmd_fit, probs= c(0.025, 0.5, .975), digits=3)

# make sure the chains converged
traceplot(hmd_fit)

stan_hist(hmd_fit, pars='mode', fill='coral', color='darkgray')
stan_hist(hmd_fit, pars='beta', fill='lightblue', color='darkgray')
```


# Numident Data

```{r}
numident <- read_csv(here("stan/data/censoc_numident_demo_v1/censoc_numident_demo_v1.csv")) 

## filter to include only one birth year 
numident <- numident %>% 
  filter(byear == 1918) %>% 
  filter(sex == "Male") 
  
## numident data
numident_stan_data = list(x = numident$death_age, N = length(numident$death_age), L = min(numident$death_age), U = max(numident$death_age) + 1)

## plot density of deaths  
ggplot(numident) + 
  geom_histogram(aes(x = death_age),
                 color = "black",
                 binwidth = 1) + 
  theme_classic(15) + 
  scale_fill_manual(values = c("grey", "red")) + 
  labs(x = "Death Age",
       y = "Density")

## look at STAN object
numident_fit =  sampling(object = simple_gomp_model, 
                 data = numident_stan_data, 
                 init = init_gomp,
                 chains = 1,
                 iter   = 1000, 
                 warmup = 200)

## 
mode <- mean(rstan::extract(numident_fit)$mode)
beta <- mean(rstan::extract(numident_fit)$beta)

numident_fit
```

## Numident data MLE

This code is written by the MLE group (Jordan, Serge, Nathan). I have made no edits. 

```{r}
dt1 <- numident[sex == "Male" & dyear %in% 1988:2005 & byear %in% 1918] ## restrict the sample to single birthyear
table(dt1$death_age, exclude = F) ## call freq table to estimate value for M.guess
hist(dt1$death_age)  ## note that death_age values are floored integers

n = 1436 ## number of males in sample
M.guess = 79 ## based on visual inspection of freq table and histogram
beta.guess = 0.1 ## beta is typically around 0.1
u = 88 ## upper bound, based on n = 1436  # upper bound needs to be one integer higher than observed max age at death
l = 69 ## lower bound, based on n = 1436
y <- dt1[, death_age] ## here we use actual ages at death instead of simulated ages
yt = y ## actual ages are already truncated, so y is also yt


## 2. MLE estimation with known truncation

minusLogLik_1 =  function(yt, p, u, l) ## p is a vector of M and log(beta)
{
    M = exp(p[1])
    beta = exp(p[2])
    n = length(yt)
    ## denom : F(u) - F(l)
    denom = pgomp_mode(u, M = M, beta = beta) - pgomp_mode(l, M = M, beta = beta)
    ## L = f1/denom * f2/denom ... = f1*f2*f3 / denom^n
    ## LL = sum(log(fi)) - n * log(denom)
    LL = sum(dgomp_mode(yt, M = M, beta = beta, log = TRUE)) - n * log(denom)
    mLL = -LL
    return(mLL)
}


## 3. Maximize the likelihood and find Hessian estimated standard errors

## starting values
p.start = c(log.M = log(M.guess),  # not multiplied by anything because doing so seems to break the function
            log.beta = log(beta.guess * 1.2))

## optimizer (to find the smallest minusLogLik_1 using p.start parameters)
fit1 = optim(par = p.start, fn = minusLogLik_1, yt = yt, u = u, l = l)

## get out estimates
log.est = fit1$par
est = round(exp(log.est),3)
names(est) = c("M.hat", "beta.hat")
guess <- c("M.guess" = M.guess, "beta.guess" = beta.guess)
print(cbind(est, guess))

## Hessian estimated std err
set.seed(123)
fit1.hess = optim(par = p.start, fn = minusLogLik_1, yt = yt, u = u, l = l,
                  hessian = TRUE)
fit <- fit1.hess

## fisher_info<-solve(-fit$hessian)
H = fit$hessian
fisher_info = solve(H)
sigma.hat <- sqrt(diag(fisher_info))
upper <- fit$par + 1.96 * sigma.hat
lower <- fit$par - 1.96 * sigma.hat
interval <- data.frame(value = fit$par, upper = upper, lower = lower)
round(interval, digits = 3)
round(exp(interval[1:2,]), digits = 3)
```

## visualize 
```{r}
## mle results
mode_mle <- est[[1]]
beta_mle <- est[[2]]
alpha_mle <- getAlpha(M=mode_mle, beta=beta_mle)

## stan results
mode_stan <- mode
beta_stan <- beta
alpha_stan <- getAlpha(M = mode_stan, beta = beta_stan)

# visualization (pretty similar)
tibble(death_ages) %>% 
  mutate(window = case_when(
    death_ages > truncation_lower & death_ages < truncation_upper ~ "Y",
    TRUE ~ "N"
  )) %>% 
ggplot(aes(x = death_ages, fill = window)) + 
  stat_function(fun=function(x) alpha_mle * exp(beta_mle*x) * exp((-alpha_mle/beta_mle) * (exp(beta_mle*x)-1)), color = "blue", size = 1) +
  stat_function(fun=function(x) alpha_stan * exp(beta_stan*x) * exp((-alpha_stan/beta_stan) * (exp(beta_stan*x)-1)), color = "red", size = 1) +
  theme_classic(15) + 
  scale_fill_manual(values = c("grey", "red")) + 
  labs(x = "Death Age",
       y = "Density") + 
  theme(legend.position="none")

tibble(method = c("mle", "stan"),
       mode  = c(mode_mle, mode_stan),
       mle = c(beta_mle, beta_stan)) %>% 
  knitr::kable()
```



```{r}
test <- numident %>% 
  # filter(incwage > 0 & incwage < 6000 & sex == "Male") %>% 
  filter(byear %in% 1915:1920)

model <- lm(death_age ~ as.factor(sex) + as.factor(byear), data = test)

summary(model) 
```

